---
title: "Introduction to the Model"
author: "Brockton Butcher"
date: "1/17/2019"
output: github_document
---

The goal of this study is to use visual characteristics gathered from a Bull and a Cow, and create a predictive model that provides a probability for the corresponding distributions of the offspring. Once this initial model is created I would like to also explore the impacts of how the grandparents/lineage of the offspring also effects probability/ strength in priors and if this could be useful in an estimate.



```{r}
d = read.csv('cowswepd')

model = "
model {
for (j in 1:J ) {
y[,j] ~ dmulti(p[,j], n[j])
p[1,j] <-   pnorm(0,       mu[j], 1.0)
p[2,j] <-   pnorm(alpha[1],mu[j], 1.0) - pnorm(0,       mu[j], 1.0)
p[3,j] <-   pnorm(alpha[2],mu[j], 1.0) - pnorm(alpha[1],mu[j], 1.0)
p[4,j] <-   pnorm(alpha[3],mu[j], 1.0) - pnorm(alpha[2],mu[j], 1.0) 
p[5,j] <- 1-pnorm(alpha[3],mu[j], 1.0)
  
  mu[j] ~ dnorm(theta,tau)

                }
for(r in 1:3){
alpha0[r]~dunif(0,10)
}
alpha <- sort(alpha0) #JAGS sort trick

# Hyperparameters for normal means
 theta ~dnorm(0,1e-6)
 tau ~ dgamma(1,1)
}
"

library(rjags)
dat=list(y=t(d[,-1]),n=rowSums(d[,-1]), J=nrow(d))
m=jags.model(textConnection(model),dat, inits=list(mu=rep(1,dat$J), alpha0=1:3))
```




```{r}
model {
  for (i in 1:Ntotal) {
    y[i] ~dcat(pr[i,1:nYlevels])
    pr[i,1] <- pnorm(thresh[1], mu[x[i]], 1/sigma[x[i]]^2)
    for (k in 2:(nYlevels-1))n {
      pr[i,k] <- max (0, pnorm(thresh[k], mu[x[i]],1/sigma[x[i]]^2 )
                      -pnorm(thresh[k-1], mu[x[i]],1/sigma[x[i]]^2 ) )
    }
    pr[i,nYlevels] <- 1-pnorm(thresh[nYlevels-1],mu[x[i]],1/sigma[x[i]]^2 ) 
  }
  for (j in 1:2 ) { #2 groups
    mu[j] ~dnorm( (1+nYlevels/2),1/(nYlevels)^2 )
    sigma[j]~dunif(nYlevels/1000, nYlevels*5)
  }
  for ( k in 2:(nYlevels-2) ) {#1 and nYlevels are fixed, not stochastic 
    thresh[k]~dnorm(k+.5, 1/2^2)
  }
}
```


#other specification
Jags-Yord-Xnom2grp-MnormalHet.R
Jags-Yord-Xnom2grp-MnormalHet-Example.R




```{r}
#priors for bulls

#read mother's score
#obtain her records
#records of siblings
#record of sire
#record of sire's offspring
#other scores in lineage

#we want to see if priors are actually a model that can be trusted with resonable confidence based on no progeny for a bull when trying to predict


```

We want to determine the prior for each bull based on his appearance and the information of the mother and the traits that she possesses, along with other performance records within his lineage. If we see that most of the time his mother has offspring with a 4 foot score, and his father has offspring mostly with a 3 footscore we would like to think that on average he would produce mostly 3's and 4's. Then through the course of time we would 
